Identifying Inter-Genomic Repeats Using Fast, Approximate Methods for Betweenness Centrality

### 2015-05-19 manuscript/manuscript.tex
Added back previous edits.

The assembly of metagenomics data, the sequences of DNA recovered from an environment without culturing, is considerably more complex than that of individual genomes.
%Challenges include: (i) repetitive sequences within individual genomes; (ii) varying abundances of different organisms in the community; (iii) variation between closely related organisms; and (iv) conserved genomic segments shared by distantly related organisms.
Repetitive sequences pose a major challenge when assembling isolated genomes.
It has been shown that the number of reconstructions grow exponentially with the number of repeats making it computationally infeasible to find one correct solution for assembly\cite{kingsford}.
The effect of repeats on assembly is exacerbated in metagenomic datasets due to the presence of conserved regions of the genome linking together distantly related organisms.
To tackle the problem of identifying these inter-genomic repeats, several assemblers have used the notion of network centrality to find repeats within the assembly graph.
For example, Bambus2\cite{bambus} uses betweenness centrality to find \emph{central} nodes in the graph.
The drawback of their approach is the expensive calculation of betweenness centrality by computing all-pair shortest paths.
This computation is impractical when applied to the large assembly graphs typically encountered.
To overcome this drawback, we explore the application of approximate and parallel betweenness centrality algorithms on metagenomic assembly graphs and evaluate their effectiveness in identifying inter-genomic repeats.

\begin{figure*}[htbp]
\centering
\includegraphics[width=0.60\textwidth]{es_mix_200kb_k21.png}
\caption{Assembly graph of a simulated community consisting of 200 Kbp subsets of \textit{Escherichia coli} str. K-12 MG1655 and \textit{Staphylococcus aureus}.  Nodes are colored and sized based on their relative betweenness centrality. The highlighted nodes are inter-genomic repeats whose deletion would separate the graph. Note that the betweenness centrality measure correctly identifies these nodes.}
\label{fig:sampled_nodes}
\end{figure*}

>>>>

The assembly of metagenomics data, the sequences of DNA recovered from an environment without culturing, is considerably more complex than that of individual genomes.
%Challenges include: (i) repetitive sequences within individual genomes; (ii) varying abundances of different organisms in the community; (iii) variation between closely related organisms; and (iv) conserved genomic segments shared by distantly related organisms.
Repetitive sequences pose a major challenge when assembling isolated genomes.
It has been shown that the number of reconstructions grow exponentially with the number of repeats making it computationally infeasible to find one correct solution for assembly\cite{kingsford}.
The effect of repeats on assembly is exacerbated in metagenomic datasets due to the presence of conserved regions of the genome linking together distantly related organisms.
To tackle the problem of identifying these inter-genomic repeats, several assemblers have used the notion of network centrality to find repeats within the assembly graph (Figure \ref{fig:es_mix}).
For example, Bambus2\cite{bambus} uses betweenness centrality to find \emph{central} nodes in the graph.
The drawback of their approach is the expensive calculation of betweenness centrality by computing all-pair shortest paths.
This computation is impractical when applied to the large assembly graphs typically encountered.
To overcome this drawback, we explore the application of approximate and parallel betweenness centrality algorithms on metagenomic assembly graphs and evaluate their effectiveness in identifying inter-genomic repeats.

\begin{figure*}[htbp]
\centering
\includegraphics[width=0.70\textwidth]{es_mix_200kb_k21.png}
\caption{Assembly graph of a simulated community consisting of 200 Kbp subsets of \textit{Escherichia coli} str. K-12 MG1655 and \textit{Staphylococcus aureus}.  Nodes are colored and sized based on their relative betweenness centrality. The highlighted nodes are inter-genomic repeats whose deletion would separate the graph. Note that the betweenness centrality measure correctly identifies these nodes.}
\label{fig:sampled_nodes}
\end{figure*}
